---
title: "Chapter 2 homework"
author: "Judah Altman"
format:
  html:
    toc: true
    toc-depth: 3
embed-resources: true
execute:
  echo: true
  message: false
  warning: false
---

## Setup

Assume you are working in an RStudio Project. Use `here::here("data", "filename.csv")` to build file paths to datasets in the `data` folder.

```{r}
library(tidyverse)
library(here)
library(psych)   # used for describe()/describeBy() in a few places
```

```{r}
heart <- read_csv(here::here("data", "heart.csv"))
```

::: callout-tip
If you ever get a “file not found” error, confirm (1) you opened the correct `.Rproj`, and (2) the dataset is inside your project’s data folder (accessed via `here::here("data", ...)`).
:::

## Tutorial (Chapter 2, plus a bit of 3)

### 1) Estimates of variation around the mean

In class, we talked about quantifying variation. Variance and standard deviation are both based on the **sum of squared error (SSE)** around a model’s predictions.

Start by computing the sample standard deviation and sample variance of height:

```{r}
# Compute sd and variance of height
sd(heart$Height)
var(heart$Height)
```

#### Convert variance to standard deviation

Recall: $\text{SD} = \sqrt{\text{Var}}$.

```{r}
# Convert variance to SD using R as a calculator
sqrt(var(heart$Height))
```

#### Compute SSE from the variance and sample size

For the *sample* variance, \$\$ s\^2 = \frac{\text{SSE}}{n-1} \quad\Rightarrow\quad \text{SSE} = s\^2 (n-1).

```{r}
# Compute n, variance, and then SSE = var * (n - 1)
var(heart$Height) * (length(heart$Height) - 1)
```

::: callout-tip
In R, `var(x)` uses the sample variance with denominator $n-1$. That’s why the formula above uses $n-1$.
:::

### 2) Estimates of central tendency

Compute mean and median:

```{r}
# Mean and median of height
median(heart$Height)
mean(heart$Height)
```

#### Mode (custom helper)

R’s base `mode()` is **not** the statistical mode. Below is a simple helper that returns the most frequent value(s). If the data are multimodal, it returns the average of the modes.

```{r}
mymode <- function(x) {
  x2 <- na.omit(x)
  ux <- unique(x2)
  tab <- tabulate(match(x2, ux))
  mean(ux[tab == max(tab)])
}
```

```{r}
# Use mymode() to compute the mode of height
mymode(heart$Height)
```

### 3) Group and summarize a dataset

Compute mean and SD of height by gender using tidyverse verbs:

```{r}
# Summarize Height by Gender:
# - n
# - mean height
# - sd height
heart |> 
  summarise(n())
heart |> 
  summarise(mean(Height))
heart |> 
  summarise(sd(Height))
```

If you want a richer descriptive table, `psych::describeBy()` can do that:

```{r}
describeBy(heart$Height, heart$Gender, mat = TRUE)
```

::: callout-tip
Even when you use a helper like `describeBy()`, prefer doing **wrangling** (filtering, selecting, mutating, grouping) with tidyverse verbs first, then pass the result to the summary function.
:::

### 4) Estimates of variation revisited: SSE from the data

One way to compute SSE around the **mean model** (predict $\bar{Y}$ for everyone):

```{r}
# Compute SSE around the mean
heart_height <- heart |>
  select(Height) |> 
  mutate(mean_pred = mean(Height),
         mean_err = Height - mean_pred)

sse_mean <- sum(heart_height$mean_err^2)
sse_mean
```

Now modify that idea to compute SSE around the **median model** (predict $\tilde{Y}$ for everyone).

```{r}
# Compute SSE around the median
heart_height <- heart_height |>
  mutate(med_pred = median(Height),
         med_err = Height - med_pred)

sse_med <- sum(heart_height$med_err^2)
sse_med
```

Answer in words: is SSE around the median larger or smaller than SSE around the mean for these data?

SSE around the median is larger than SSE around the mean for this data.

### 5) Missing data

Many functions accept `na.rm = TRUE` to ignore missing values.

```{r}
# Example: mean height ignoring missing values
mean(heart$Height, na.rm = TRUE)
```

You can also drop missing values explicitly:

```{r}
heart |> 
  drop_na(Height) |> 
  summarize(mean_height = mean(Height))
```

## HW 02 Questions

### 1) Central tendency and variability (USNEWS)

The dataset `USNEWS.csv` contains data used by *U.S. News and World Report* to make its college rankings. Two variables of interest are:

-   `gradRate`: graduation rate (percent from 0 to 100)
-   `accptRate`: acceptance rate (proportion from 0 to 1)

For each variable, obtain estimates of central tendency (mean, median, mode) and variability (variance and standard deviation). Write a few sentences describing what you found.

```{r}
usnews <- read_csv(here::here("data", "USNEWS.csv"), na = c(".", "NA"))
?read_csv
```

```{r}
# Create a compact summary table for gradRate and accptRate:
# mean, median, mode, variance, sd, and n (non-missing)
grad_Rt <- usnews$gradRate[!is.na(usnews$gradRate)]
accpt_Rt <- usnews$accptRate[!is.na(usnews$accptRate)]
summary_tbl1 <- as.tibble(describeBy(grad_Rt))
summary_tbl2 <- as.tibble(describeBy(accpt_Rt)) 
summary_tbl <- bind_rows(summary_tbl1, summary_tbl2) |> 
  select(n:median) |> 
  mutate(mode = c(mymode(grad_Rt), mymode(accpt_Rt)),
         var = c(var(grad_Rt), var(accpt_Rt)))
summary_tbl
```

For this data, the mean, median and mode are all relatively similar for graduation rate, but for acceptance rate, while the median and mean are similar, the mode is 1, meaning that the most common acceptance rate among schools is 100%. For standard deviation and variance, the most interesting finding is that the graduation rate standard deviation unit is around 18.9, meaning that the variance is around 357, which implies that the data is very spread around the range of values from 0 to 100, rather than concentrating around the mean.

### 2) Conditional vs unconditional predictions (USNEWS)

Another interesting variable is `Type` (public vs private).

1.  Write **Model C** that makes a constant prediction for every school.
2.  Write **Model A** that makes predictions of `gradRate` conditional on `Type`.
3.  As a first look at whether Model A might be useful, compute the mean `gradRate` for private and public schools.
4.  Write a sentence or two describing these results and whether it *appears* useful to move from Model C to Model A.

Use both a verbal description and a model statement in LaTeX.

#### Model statements (fill in)

-   Model C (compact): $$
    Y_i = b_0 + \varepsilon_i
    $$

Here, $Y_i$ is Yth observation of the data, $b_0$ is a constant prediction of the graduation rate, and $\varepsilon_i$ is the error of the prediction from the observation.

ex. $$ GraduationRate_i = 50 + \varepsilon_i $$

-   Model A (augmented; conditional means by type): $$
    Y_i = b_0 + b_1 X_i + \varepsilon_i
    $$ where $X_i$ is an indicator you define (e.g., $X_i = 1$ if Private, $0$ if Public).

Here, the same as above is true, but $X_i$ is the type of school (private or public) and $b_1$ is the predicted change in graduation rate based on the type of school.

ex. $$ GraduationRate_i = 50 + 15 X_i + \varepsilon_i $$

```{r}
# Compute mean gradRate by Type (and include n)
usnews |> 
  group_by(Type) |> 
  summarize(mean(gradRate, na.rm = TRUE))
```

Based only on the mean of the graduation rate by type of school, it is likely useful to move from model C to model A because the mean graduation rate differs by 16 percentage points from private to public, so whether or not a school is private or public likely affects what the graduation rate of that school will be, therefore the prediction will likely be more accurate if it incorporates a parameter for the type of school.

### 3) Coupon campaign and store sales (stores)

At 10 grocery stores, a market researcher records the number of cases of a product sold both before and after coupons were mailed to households in the area. The data below are the **changes** in the number of cases sold (positive = more after coupons, negative = fewer, zero = no change).

| Store | Change |
|------:|-------:|
|     A |      5 |
|     B |      4 |
|     C |     -2 |
|     D |      6 |
|     E |      1 |
|     F |      0 |
|     G |     -4 |
|     H |      3 |
|     I |      2 |
|     J |      7 |

#### 3a) By hand

Calculate the **mean** and **standard deviation** for the change scores by hand (or in Excel).

The mean of the change scores is 2.2 and the standard deviation is \~3.5

(Optional check in R after you finish your by-hand work:)

```{r}
# Create a vector of the change scores and compute mean/sd as a check
change <- c(5, 4, -2, 6, 1, 0, -4, 3, 2, 7)
mean(change)
sqrt(sum((mean(change) - change)^2) / 9 ) #a bit of a pain to calculate by hand
sd(change)
```

#### 3b) Model C (no change)

Specify a Model C that predicts **no change** for every store.

Write it in the same form as class:

$$
Y_i = \text{(prediction)} + \varepsilon_i
$$ $$
Change_i = 3 + \varepsilon_i
$$

```{r}
# Write Model C in words (in the text), then compute SSE(C) in R later
model_c_err <- change - 3 
```

#### 3c) Model A (best constant prediction from the data)

Specify a Model A with one parameter that makes the best possible constant prediction of `Change` based on the data.

$$
Y_i = b_0 + \varepsilon_i
$$ $$
Change_i = Change_\mu + \varepsilon_i
$$ where $Change_\mu$ is the mean value of the change scores.

```{r}
# Write Model A in words (in the text), then compute SSE(A) in R later
model_a_err <- change - mean(change) 
```

#### 3d) Null hypothesis and interpretation

State the null hypothesis tested by comparing Model C and Model A, and give a non-technical interpretation of what the comparison asks.

The null hypothesis tested by comparing Model C and Model A is that Model A does not meaningfully reduce the error in Model C and that either there is no difference in error between Model C and Model A or that the reduction in error does not warrant accepting Model A over Model C. This comparison asks which Model best represents the data in the most parsimonious fashion, and accepting the null hypothesis means Model A does not improve on Model C.

#### 3e) Compute error for both models (SSE)

Using SSE as your aggregate measure of error, compute:

-   $\text{SSE}(C)$ for the compact model
-   $\text{SSE}(A)$ for the augmented model

```{r}
# Compute SSE(C) and SSE(A)
sse_a <- sum(model_a_err^2)
sse_a
sse_c <- sum(model_c_err^2)
sse_c
```

#### 3f) Proportional reduction in error (PRE)

Compute: $$
\text{PRE} = \frac{\text{SSE}(C) - \text{SSE}(A)}{\text{SSE}(C)}.
$$

Based on this PRE, do you think Model C should be rejected in favor of Model A? (No formal test yet—explain your reasoning.)

```{r}
# Compute PRE from SSE(C) and SSE(A)
pre <- (sse_c - sse_a) / sse_c
pre 
```

Based on this PRE and the small number of values, I think Model C should not be rejected in favor of Model A because it only reduces the error by around 5% on a very small sample size.

#### 3g) Do it again using `stores.csv`

These data are also available in `stores.csv`. Read in the data and use R to obtain all the numbers you calculated above (including PRE if you can).

```{r}
stores <- read_csv(here::here("data", "stores.csv"))
```

```{r}
# Use the stores data to compute:
# - mean and sd of Change
ch_mean <- mean(stores$Change)
ch_mean
sd(stores$Change)

# - SSE(C) where prediction is 0
sse_c <- sum((stores$Change - 0)^2)
sse_c

# - SSE(A) where prediction is mean(Change)
sse_a <- sum((stores$Change - mean(ch_mean))^2)
sse_a

# - PRE
pre <- (sse_c - sse_a) / sse_c
pre 
```

::: callout-tip
A convenient SSE pattern is:

-   If predictions are stored in `y_hat`, then `sum((y - y_hat)^2)`.
-   For Model C here, `y_hat` is a constant 0.
-   For Model A here, `y_hat` is a constant equal to `mean(y)`.
:::

### 4) Concept questions

In your own words (**not** using the example from class):

1.  Why is $\text{ERROR}(\text{Model A}) \le \text{ERROR}(\text{Model C})$?

Error(Model A) will always be less than or equal to the Error(Model C) because adding parameters will always slightly reduce the error even if it is not a meaningful addition that reduces the error enough that it is worth adding to the model.

2.  What is a **degree of freedom**?

Degrees of freedom is the degree to which values are "free" to take on any value. Therefore a degree of freedom of 1 means that there is only one value that an unknown value can take on. Adding parameters to a model reduces the degree of freedom for each parameter added from the total number of values in the data.

### 5) Sampling distributions and small samples

Visit the app:

-   https://correll.shinyapps.io/centralTendency/

5a) Generate four or five iterations using a normally distributed population and samples of $n = 5$.

-   Where do the sampling distributions of the mean peak?
    -   The sampling distributions of the mean tend to peak around the middle (which is 10 here)
-   How much do the means and medians vary? What minimum and maximum values do you see for each?
    -   The means and medians don't vary too much in that they are both centered around 10, but the mean is narrower (narrower distribution around the meanb), and the median is wider (wider distribution around the median)

5b) Answer the same questions for the mean using $n = 500$. (Rescale the histograms so you can see variability.)

-   Where do the sampling distributions of the mean peak?
    -   Similarly around 10 for both, but all are much closer to 10 this time.
-   How much do the means and medians vary? What minimum and maximum values do you see for each?
    -   The two values differ much less here although again the median has a slightly wider distribution around 10.

### 6) Central Limit Theorem intuition

Visit the app:

-   https://correll.shinyapps.io/centralLimit/

6a) Set a **skewed** population and draw samples of different sizes. What is the shape of the sampling distribution of the mean for samples of $n = 2$?

-   For a skewed population of n = 2, the shape of the sampling distribution of the mean is all concentrated between 0 and 10, with most values equal to around 3 to 4.

6b) What is the shape for $n = 10$?

-   For n = 10, it is concentrated between 0 and around 5, with most values equal to around 4.

6c) What is the shape for $n = 100$?

-   For n = 100, it is concentrated around between 3 and 5, with most values equal to around 4.
