---
title: "Chapter 4 homework"
author: "Judah Altman"
format: html
embed-resources: true
---

```{r}
#|message: false

library(tidyverse)
library(broom)
library(here)
```

```{r}
fitness <- read_csv(here("data", "fitness.csv"))
```

# Homework questions

## 1. Sampling distributions and statistical surprise

Write brief answers.

-   What is a sampling distribution of a statistic (like a mean), and what is it useful for?

A sampling distribution is the theoretical amount you approach if you sample an infinite number of times. In the case of a sampling distribution of a statistic like the mean, it is the distribution of that statistic taken from many samples.

-   How can a sampling distribution be used to evaluate whether an observed value is “surprising” under a null hypothesis?

It can be used by looking at the sampling distribution of the null hypothesis many times and seeing if the value that is observed falls within that range. If the likelihood of that value falling in the sampling distribution is extremely unlikely, then it is a surprising as compared to the null hypothesis.

## 2. Model comparison test: is the mean resting pulse 72?

Using the Fitness data, evaluate whether the sample of participants is consistent with a population mean of `RSTPULSE = 72` using a **model comparison**.

### 2a. Fit Model C and Model A

Use these models:

-   **Model C (null):** the outcome is fixed at 72 with no free parameters\
    $$Y_i = 72 + \varepsilon_i.$$

-   **Model A (mean model):** estimate the mean from the data\
    $$Y_i = b_0 + \varepsilon_i.$$

```{r}
# Model C: fixed mean at 72 (no parameters)
# Hint: use lm(... ~ 0) with the outcome shifted by 72.
fitness <- fitness |> 
  mutate(rstpulse_dev = RSTPULSE - 72)

model_c <- lm(rstpulse_dev ~ 0, data = fitness)

# Model A: estimate the mean (intercept-only model)
model_a <- lm(rstpulse_dev ~ 1, data = fitness)
```

### 2b. Compute PRE and F

Compute:

$$\text{SSE}(\cdot) = \sum_i e_i^2$$ (use `deviance()` for an `lm` object)

$$\text{PRE} = \frac{\text{SSE}(C) - \text{SSE}(A)}{\text{SSE}(C)}$$

$$F^* = \frac{\text{PRE}/(P_A - P_C)}{(1-\text{PRE})/(n - P_A)}$$

```{r}
sse_c <- deviance(model_c)
sse_a <- deviance(model_a)

pre <- (sse_c - sse_a) / sse_c
pre

n <- nrow(fitness)
pc <- 0
pa <- 1

f_star <- (pre / (pa - pc)) / ((1 - pre) / (n - pa))
f_star
```

### 2c. Identify “surprising” values from the book tables and/or by using `qf()`.

Fill in the quantities the tables depend on:

-   number of participants: n = 31
-   number of parameters in Model C: $P_C$ = 0
-   number of parameters in Model A: $P_A$ = 1
-   new parameters: $P_A - P_C$ = 1
-   unused parameters (residual df for Model A): $n - P_A$ = 30

Then, using the book tables and/or `qf()`:

-   critical PRE at $\alpha = .05$: PREcrit = 0.122
-   critical F at $\alpha = .05$: Fcrit = 4.17

### 2d. Conclusions

-   Statistical conclusion: Do you reject or retain the null hypothesis at $\alpha = .05$? Explain using the logic of “surprising under the null.”

You should reject the null hypothesis at $\alpha = .05$ because the F-statistic exceeds its critical value, and is therefore "surprising". In other words, the error reduction by adopting the parameter in model A over model C is much greater than the average reduction in error that could be achieved by just adding a random parameter.

-   Substantive conclusion: Give a one-sentence interpretation in plain language.

In the sample, the resting pulse of individuals was better predicted by the average resting pulse of the sample rather than the population average, and therefore, this sample's average resting pulse was significantly different from the population's average resting pulse of 72.

## 3. Alternative F formula using SSR and SSE(A)

Let

$$\text{SSR} = \text{SSE}(C) - \text{SSE}(A)$$

$$F^* = \frac{\text{SSR}/(P_A - P_C)}{\text{SSE}(A)/(n - P_A)}$$

### 3a. Compute F using SSR and SSE(A)

```{r}
ssr <- sse_c - sse_a
ssr
f_star_ssr <- (ssr / (pa - pc)) / (sse_a / (n - pa))
f_star_ssr
```

### 3b–3e. Interpretation

Write brief answers.

-   What does the numerator $\text{SSR}/(P_A - P_C)$ represent?

The numerator here represents mean squares reduced.

-   What does the denominator $\text{SSE}(A)/(n - P_A)$ represent?

The denominator here represents the mean square error, or the variance.

-   What does $F^*$ represent conceptually?

<!-- -->

-   $F^*$ represents the proportional reduction in error per additional parameter added, in relation to the number of parameters that could have been added.

<!-- -->

-   Compute $t$ for this $F^*$ (use the relationship between $t$ and $F$ when the numerator df is 1).

```{r}
t_star <- sqrt(f_star)
t_star
```

## 4. Does running increase pulse rate?

Using the Fitness dataset, compare `RSTPULSE` (resting pulse) to `RUNPULSE` (post-run pulse).

### 4a. Set up Model C and Model A

Use a model comparison where the null corresponds to **no mean change** from resting to post-run.

One convenient approach:

1.  Create a difference score $D_i = \text{RUNPULSE}_i - \text{RSTPULSE}_i$.
2.  Compare:
    -   **Model C:** $D_i = 0 + \varepsilon_i$ (no free parameters)\
        $$D_i = 0 + \varepsilon_i.$$
    -   **Model A:** $D_i = b_0 + \varepsilon_i$ (estimate the mean difference)\
        $$D_i = b_0 + \varepsilon_i.$$

```{r}
fitness <- fitness |>
  mutate(d = RUNPULSE - RSTPULSE)

model_c <- lm(d ~ 0, data = fitness)
model_a <- lm(d ~ 1, data = fitness)
```

### 4b. Compute PRE and F, then compare to critical values

```{r}
sse_c <- deviance(model_c)
sse_a <- deviance(model_a)
pre <- (sse_c - sse_a) / sse_c
pre
f_star <- (pre / 1) / ((1 - pre) / (n - 1))
f_star
```

From the book tables at $\alpha=.05$:

-   PREcrit = 0.122
-   Fcrit = 4.17

### 4c. Conclusions

-   Statistical conclusion (reject/retain the null).

Reject the null hypothesis.

-   Substantive conclusion (plain-language interpretation).

The average pulse after running is significantly higher than it is when resting. According to this model from the sample, running increases pulse rate.

## 5. With your own data

Return to your own dataset and the model comparison you set up previously. In this section, please use

```{r}
gss2024 <- readRDS(here("data", "gss2024.rds"))

df <- gss2024 |> 
  select(eqwlth) |> 
  drop_na() |> 
  haven::zap_labels() 
```

### 5a. Fit your models

```{r}
df <- df |>
  mutate(eqwlth_dev = eqwlth - 4)

model_c <- lm(eqwlth_dev ~ 0, data = df)
model_a <- lm(eqwlth_dev ~ 1, data = df)
```

### 5b. Report PRE, F\*, and t\*

```{r}
pre <- (deviance(model_c) - deviance(model_a)) / deviance(model_c)
pre
f_star <- (pre / 1) / ((1 - pre) / (nrow(df) - 1))
f_star
t_star <- sqrt(f_star)
t_star
```

### 5c. Use tables or functions to identify critical values and write conclusions

```{r}
f_crit <- qf(0.95, 1, 2160)
f_crit
f_crit < f_star
```

The critical value of F in this case is \~3.8 and the F\* value of \~165 indicates that we should reject the null hypothesis that the average individual will indicate a 4 (that the government should be somewhat involved and somewhat not involved in reducing income differences), and that the mean of this value is a better predictor of what the sample thinks the government's involvement in reducing income should be.