---
title: "Homework: Chapter 8 and Model Comparison"
format: html
---

## Setup

Use the same `states` data from the chapter, with the `inc1000` variable.

```{r}
pacman::p_load(dplyr, tidyr, broom, stringr, modelsummary, tinyplot, ggplot2)

state.x77_with_names <- as_tibble(state.x77, rownames = "state")

states <- bind_cols(state.x77_with_names,
                    region = state.division) |>
  janitor::clean_names() |>
  mutate(south = as.integer(str_detect(as.character(region), "South")),
         inc1000 = income / 1000)
```

## The Additive Model

Estimate this additive model and display the results with `msummary()`:

```{r}
m1 <- lm(life_exp ~ hs_grad + inc1000, data = states)

msummary(list("M1" = m1),
         estimate = "{estimate} ({std.error})",
         statistic = NULL,
         gof_map = c("nobs", "F", "r.squared"))
```

Write a sentence interpreting each of the two slope coefficients. Use appropriately hedged language (e.g., "adjusting for...").

1\) Adjusting for income, for every one additional percent increase in high school graduation, life expectancy by state is expected to increase by 0.1 years according to the model.

2\) Adjusting for high school graduation, for every increase in 1000 in the average income of a state, life expectancy by state is expected to decrease by 0.073 years.

Now compute and examine the 95% confidence interval for `inc1000`. What does it tell you about the precision of that estimate? What might this imply about whether income is a useful predictor of life expectancy after accounting for education?

```{r}
#'by hand' way
f_crit <- qf(0.95, 1, 47)
bounds <- sqrt((f_crit * (deviance(m1) / (nrow(states) - 3))) / ((nrow(states) - 1) * var(states$inc1000) * (1 - summary(m1)$r.squared)))
lower_bound <- m1$coefficients["inc1000"] - bounds
upper_bound <- m1$coefficients["inc1000"] + bounds
ci <- c(lower_bound, upper_bound)
ci

#fast way (close, but not quite the same... ask about this in class)
tidy(m1, conf.int = TRUE, conf.level = .95)
```

It is an imprecise estimate. It implies that income is not a useful predictor of life expectancy after accounting for education because it the variation is too high to accurately predict life expectancy based on income.

## Visualizing the Additive Model

The companion shows two ways to plot `m1`: with `hs_grad` on the x-axis at representative values of income, and then with axes swapped. Make one of these plots (your choice).

```{r}
myincs <- quantile(states$inc1000, c(0, 0.5, 1))
myhs <- seq(min(states$hs_grad), max(states$hs_grad), length = 100)
xgrid <- expand_grid(inc1000 = myincs, hs_grad = myhs)

#predictions
m1_yhat <- predict(m1,
                newdata = xgrid,
                interval = "confidence",
                level = .95)

m1_preds <- bind_cols(m1_yhat, xgrid) |>
  mutate(income = case_match(inc1000,
                             min(states$inc1000) ~ "min",
                             max(states$inc1000) ~ "max",
                             .default = "median"))

ggplot(m1_preds, aes(x = hs_grad, y = fit, color = income, fill = income)) +
  geom_ribbon(aes(ymin = lwr, ymax = upr), alpha = 0.2, color = NA) +
  geom_line(linewidth = 0.5) +
  labs(
    title = "Predicted life expectancy",
    subtitle = "Additive model, 95% CIs",
    y = "Life expectancy",
    x = "% of HS grad",
    color = "income",
    fill = "income") +
  theme_minimal()
```

In an additive model, the prediction lines for different values of the other variable are always parallel. Explain in 2--3 sentences what the parallelism means substantively. What assumption is the additive model making about how `hs_grad` and `inc1000` jointly affect life expectancy?

The prediction lines are parallel in the additive model because the contribution of each predictor to the best predicted values does not depend on the values of the other predictors. Therefore, regardless of the level of income, for example, the level of high school graduation rate won't change. In other words, the slopes in each case for each predictor are independent of each other, so any values will not change the slope of the other variable, meaning the slopes will stay the same and thus be parallel.

## The Interaction Model

Now estimate:

```{r}
m2 <- lm(life_exp ~ hs_grad * inc1000, data = states)

msummary(list("M1" = m1, "M2" = m2),
         estimate = "{estimate} ({std.error})",
         statistic = NULL,
         gof_map = c("nobs", "F", "r.squared"))
```

Display `m1` and `m2` side by side using `msummary()`.

The interaction coefficient is negative. What does that mean? Describe the interaction from both directions: what does it say about the slope of `hs_grad` at different levels of income, and what does it say about the slope of `inc1000` at different levels of education?

A negative interaction coefficient means that the simple slope for graduation decreases by 0.096 as income increases by 1000 dollars, and the simple slope for income decreases by 0.096 as graduation increases by one percentage point. In other words, as income increases, or at higher levels of income, the effect of graduation on life expectancy will be less (decrease), and as graduation rate increases, or at high graduation rates, the effect of income on life expectancy will be less (decrease).

## Visualizing the Interaction

Make the interaction plot from the companion --- predicted `life_exp` against `hs_grad` at three representative values of `inc1000` (minimum, median, maximum). Use either tinyplot or ggplot2.

```{r}
#predictions with m2
m2_yhat <- predict(m2,
                newdata = xgrid,
                interval = "confidence",
                level = .95)
m2_preds <- bind_cols(m2_yhat, xgrid) |>
  mutate(income = case_match(inc1000,
                             min(states$inc1000) ~ "min",
                             max(states$inc1000) ~ "max",
                             .default = "median"))

ggplot(m2_preds, aes(x = hs_grad, y = fit, color = income, fill = income)) +
  geom_ribbon(aes(ymin = lwr, ymax = upr), alpha = 0.2, color = NA) +
  geom_line(linewidth = 0.5) +
  labs(
    title = "Predicted life expectancy",
    subtitle = "Interaction model, 95% CIs",
    y = "Life expectancy",
    x = "% of HS grad",
    color = "income",
    fill = "income") +
  theme_minimal()
```

What's the key visual difference between this plot and the additive model plot? In the footnote to this plot, the companion hints that Alaska and Utah might be driving the interaction. What would you want to investigate if this were a real analysis?

The key visual difference between this plot and the additive model plot is that for each income level, the slopes of high school graduation rates effect on life expectancy is different (intersecting), rather than parallel. This is because for high levels of income, the term interacts negatively with the high school graduation rate slope, making the slope negative. If this were a real analysis, you might want to check the same model with and without Alaska and Utah to see if the effect still holds in both cases.

## Comparing the Models

Use `anova(m1, m2)` to run the incremental F-test.

```{r}
anova(m1, m2)
```

What is the null hypothesis of this test? What do you conclude from the result?

The null hypothesis of this test is that there is no interaction effect between income and graduation rate on life expectancy. We can conclude from these results, give that the F value exceeds the critical value of F at a significance level less than 0.01 that there is a significant effect between income and graduation rate such that adding their interaction effect to the model meanignfully improves it.

Now run the likelihood ratio test with `lmtest::lrtest(m1, m2)`. The p-value from the LRT is smaller than the one from the F-test. The companion explains why --- state that explanation in your own words.

```{r}
lmtest::lrtest(m1, m2)
```

The p-value in the likelihood ration test is smaller than in the F-test because the F-test does a better job accounting for smaller sample sizes because it has a thicker tail and thus accounts for outliers better, especially in smaller sample sizes where there effect is greater. Therefore, the p-value will be greater in the F-test because the possibility for a type 1 error given those thicker tails is higher as it better accounts for the outliers that could be driving the results.

## Likelihood and Deviance

*These questions draw on the "Likelihood comparison" bonus section.*

Using `logLik()`, extract the log-likelihoods of `m1` and `m2`. Compute their deviances ($D = -2\ell$) by hand.

```{r}
ll_m1 <- logLik(m1)
ll_m1
ll_m2 <- logLik(m2)
ll_m2

deviance_m1 <- -2 * as.numeric(ll_m1)
deviance_m1
deviance_m2 <- -2 * as.numeric(ll_m2)
deviance_m2
```

The companion notes that `deviance()` on an `lm` object does *not* give you $-2\ell$ --- it gives you the RSS. Confirm this by comparing `deviance(m1)` to your hand-calculated $D$ for `m1`. Why are they different? What does each one measure?

```{r}
sse_1 <- deviance(m1)
sse_1
sse_1 == deviance_m1

sse_2 <- deviance(m2)
sse_2
sse_2 == deviance_m2
```

They are different because the sum of squared errors which is the deviance function in R, measures the sum total of each residual squared, while deviance measures the sum of log-likelihoods (how likely each data point is in the model) times -2. Therefore, deviance is not the sum of squared errors, but rather measures how far away each data point is from the model based on likelihood.

Compute the proportional reduction in deviance ($R^2$) for `m1` using both the RSS formula and the $\Delta D$ formula from the companion. Confirm they give the same answer.

```{r}
sse_c <- deviance(m1)
sse_a <- deviance(m2)

pre_r2 <- (sse_c - sse_a) / sse_c
pre_r2

dev_r2 <- 1 - exp(-(deviance_m1 - deviance_m2) / nrow(states))
dev_r2

round(pre_r2, 15) == round(dev_r2, 15)
```

## AIC and BIC

*These questions draw on the "Information criteria" bonus section.*

Estimate a null model (`m0`), the additive model (`m1`), and the interaction model (`m2`). Then compute AIC and BIC for all three.

```{r}
m0 <- lm(life_exp ~ 1, data = states)

AIC(m0, m1, m2)
BIC(m0, m1, m2)
```

Which model is preferred by AIC? Which by BIC? Are they the same?

M2 is preferred by both AIC and BIC. However, the values are not the same.

The companion explains that the AIC penalty of $2k$ corresponds to the expected increase in out-of-sample deviance per extra parameter. In your own words, what problem is the penalty term meant to solve? Why can't we just pick the model with the lowest deviance (i.e., the best in-sample fit)?

The penalty term is meant to solve the problem of overfitting. If there are too many parameters, the model will overfit to the data it is trained on, meaning that if the same model were to be tested on a new set of data, it might do a bad job of predicting the new model if it corresponds to precisely to the training data. Therefore, excess parameters have a penalty cost to make sure that the likelihood of the model only does better over and above the number of parameters that it has.

The BIC penalty is $k \log n$ instead of $2k$. For $n = 50$, compute $\log(50)$ and compare it to 2. What does this tell you about how the two criteria will differ in their preferences when $n$ is small vs. large?

```{r}
log(50)
```

log(50) \~ 3.9 which is greater than 2, which means that as n increases, the penalty term will increase. therefore, when n is small, each parameter increase will have less of an effect on deviance versus when n is large, each parameter will need to increase deviance by a greater factor to be preferred to the previous model with less parameters.
