---
title: "Class Notes Week 2"
format: html
editor: visual
---

```{r}
library(here)
library(tidyverse)
library(gssr)
library(gssrdoc)
library(broom)
```

Get data and save it to hard drive

https://gss.norc.org

```{r}
#| eval: false

gss2024 <- gss_get_yr(2024)

saveRDS(gss2024, file = here("data", "gss2024.rds"))
```

Loading saved data.

```{r}
gss2024 <- readRDS(here("data", "gss2024.rds"))
glimpse(gss2024) #gives rows, columns, and column names + data type + first few rows

df <- gss2024 |> 
  select(polviews, sex) |> 
  drop_na() |> 
  haven::zap_labels() #removes the value/cell labels (not variable/column labels), useful for information purposes, but can confuse R
```

Model C (compact model), getting R to "estimate" model C (0 parameter guess)

```{r}
df$mod_c_pred <- 4 #alternative way, for adding one quick column
df <- df |> 
  mutate(mod_c_pred = 4,
         mod_c_err = polviews - mod_c_pred)

sse_c <- sum(df$mod_c_err^2) #sum of squared errors for model C
```

Model A (augmented model)

```{r}
df <- df |> 
  mutate(mod_a_pred = mean(polviews),
         mod_a_err = polviews - mod_a_pred)

sse_a <- sum(df$mod_a_err^2)
```

Estimate proportional reduction in error (PRE)

```{r}
pre <- (sse_c - sse_a) / sse_c
```

Getting model c with lm(). Make a version of polviews "centered" on 4

```{r}
df <- df |> 
  mutate(polviews_dev = polviews - 4) #transforming 4 into 0, trick to get lm to work

mod_c <- lm(polviews_dev ~ 0, data = df) #0 is saying don't use an intercept
summary(mod_c)
deviance(mod_c) #gives SSE
```

Same for model a.

```{r}
mod_a <- lm(polviews_dev ~ 1, data = df) #1 is saying use one parameter, assume there is one intercept which 1 is multiplied by, so it estimates based on the mean (can only be 0 or 1)
summary(mod_a)
deviance(mod_a)

pre <- (deviance(mod_c) - deviance(mod_a)) / deviance(mod_c)
```

How to add a prediction to the dataframe

```{r}
df$mod_a_pred <- predict(mod_a) #predictions from the result of a model
df$mod_c_pred <- predict(mod_c)
```

Sigma is the RMSE.

```{r}
glance(mod_a)
```

New dataset, based on political views by sex.

```{r}
df2 <- df |> 
  group_by(sex) |> #creates two distributions, shifted on the x axis (slightly different means, same variance)
  summarize(mean_pv = mean(polviews))
```

Make a new variable with a meaningful 0.

```{r}
df <- df |> 
  mutate(male = ifelse(sex == 1, 1, 0))

model3 <- lm(polviews ~ 1 + male, data = df)
tidy(model3) #similar to summary()
class(model3) #lm
```

Another way is to use the factor wrapper

```{r}
model3 <- lm(polviews ~ 1 + factor(sex), data = df) #lowest value becomes the reference
tidy(model3)
```

A good way especially for plotting is to have a `chr` (character) version.

```{r}
df <- df |> 
  mutate(sex_chr = if_else(sex == 1, "Male", "Female"))

model3 <- lm(polviews ~ 1 + sex_chr, data = df)
tidy(model3)
```

Compare 1 parameter and 2 parameter model

```{r}
deviance(mod_a) #just b0
deviance(model3) #2 parameter model, b0 and b1, slight reduction in deviance

(deviance(mod_a) - deviance(model3)) / deviance(mod_a) #pre
glance(model3) #r.squared is the PRE
``` 


