---
title: "Week 3 Notes"
author: "Judah Altman"
format: html
editor: visual
---

#### Setup

```{r}
library(gssr)
library(gssrdoc)
library(tidyverse)
library(broom)
library(here)
theme_set(theme_minimal())
```

Load data

```{r}
gss2024 <- readRDS(here("data", "gss2024.rds")) |> 
  haven::zap_labels()
```

Get skewed data (this is right skewed because the tail points to the right)

```{r}
df <- gss2024 |> 
  select(tvhours) |> 
  drop_na()

ggplot(df,
       aes(x = tvhours)) +
  #geom_histogram(bins = 25, binwidth = 1, color = "white", fill = "black") #starts at 0, sets the size of the bins/cuts, color is the outline, fill is the color of the bars
  geom_bar() #bar instead of histogram because they are integer values
```

#### Sampling and visualizing

Consider this sample a population and take repeated samples from it.
First step is to write a function that grabs a sample and computes the mean.

```{r}
#this example only works on this df, to make general would need to add a df argument, also what to deal with missing values, etc.
get_sample_mean <- function(n) { 
  df |> 
    slice_sample(n = n, replace = TRUE) |> 
    summarise(m = mean(tvhours)) |> 
    as.numeric()
}
```

Then make a simulation "skeleton" that I can plug results into

```{r}
sims <- tibble(
  sim_number = 1:1000
)
```

Now I add the sampled means to the skeleton.

```{r}
sims <- sims |> 
  rowwise() |> #does the function separately by row, instead of vectorized (for the vector), saying group_by(row)
  mutate(m = get_sample_mean(n = 1)) #change n to get different distributions
```

plotting the results

```{r}
ggplot(sims, aes(x = m)) +
  geom_histogram(bins = 25, binwidth = 1, #change these as needed/desire for visualizing the results
                 fill = "gray", color = "white",
                 aes(y = after_stat(density))) +
  geom_density(color = "red")
```

#### PRE and the F-Test

Null hypothesis and null distribution: pre-data guess (often about group differences), then evaluate that guess relative to what we see in the data/sample

Null hypothesis: the average American adult in 2024 watches 3 hours of "TV" per day

```{r}
df <- df |> 
  mutate(tv_dev = tvhours - 3) #distributed around 0

m0 <- lm(tv_dev ~ 0, data = df)
m1 <- lm(tv_dev ~ 1, data = df)

observed_pre <- (deviance(m0) - deviance(m1)) / deviance(m0) #SSE_C - SSE_A / SSE_C
pre
```

Construct a null distribution - and assume that it is true

```{r}
#functions that allow you to take a sample from an infinite sample
#binom distribution (sampling from 0 or 1)
rbinom(10, 1, 0.5) #10 samples, number of trials, 50% probability
#normal distribution
rnorm(10, mean = 3, sd = 3.3) 

fake_data <- tibble(tvhours = rnorm(2152, mean = 3, sd = 3.3))

fake_data <- fake_data |> 
  mutate(tv_dev = tvhours - 3)

m0 <- lm(tv_dev ~ 0, data = fake_data)
m1 <- lm(tv_dev ~ 1, data = fake_data)

pre <- (deviance(m0) - deviance(m1)) / deviance(m0) #SSE_C - SSE_A / SSE_C
pre
```

PRE - amount that model a improves on model c even though model c is accurate, want our new model to improve more than another model would by accident

can then check how often the null distribution does not have a significant improvement through many simulations? 
Make a function

```{r}
calc_null_pre <- function() {
  null_data <- tibble(tvdev = rnorm(2152, mean = 0, sd = 3.3))
  m0 <- lm(tvdev ~ 0, data = null_data)
  m1 <- lm(tvdev ~ 1, data = null_data)
  pre <- (deviance(m0) - deviance(m1)) / deviance(m0)
  return(pre)
}
```

Create a skeleton and append the null PRE simulations, 1000 PRE simulations

```{r}
null_sims <- tibble(sim_number = 1:1000) |> 
  rowwise() |> 
  mutate(pre = calc_null_pre())
```

Show how the observed pre compares to the distribution of Null PREs - observed is quite different than the estimations, in other words, out of all the worlds we estimated, the real world (observed), is very different/unique than all the other worlds

```{r}
ggplot(null_sims,
       aes(x = pre)) +
  geom_histogram() +
  geom_vline(xintercept = observed_pre, #compares this simulation to the observed PRE
             linetype = "dashed",
             color = "red")

quantile(null_sims$pre, c(0.95, 0.99, 1)) #above the all results, 95% percentile result, 99% percentile result, and highest result
```

Based on the formula in the book, what F would that be equivalent to?

How much better did we do because of adding one parameter (numerator), over left over PRE divided by how many parameters are left (denominator)

```{r}
fstat <- (observed_pre / 1) / ((1 - observed_pre) / (nrow(df) - 1))

```

F-distribution with one numerator df and 2141 denom df

```{r}
#cf. notes
```

Can get F distributions

```{r}
?qf
qf(.99, 1, 2151) #probability, parameters added, degrees of freedom
qf(0.95, 1, 30)
```

You can get the P-value

```{r}
1 - pf(fstat, 1, 2151) #probability of getting this fstatistic (1 - probability = pvalue)
```



